Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-11-12 paper/paper.tex
start work on syntactic metatheory and denotational semantics in appendix

\newcommand{\SPF}{\mathsf{SPF}}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\K}{\mathsf{K}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\roll}{\mathsf{roll}}
\newcommand{\fold}{\mathsf{fold}}
\newcommand{\inl}{\mathsf{inl}}
\newcommand{\inr}{\mathsf{inr}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\String}{\textbf{String}}
\newcommand{\stringg}{\mathtt{String}}
\newcommand{\charg}{\mathtt{Char}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Syn}{\mathbf{Synx}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathtt{Bool}}
\newcommand{\true}{\mathtt{true}}
\newcommand{\false}{\mathtt{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
\newcommand{\theoryabbv}{$\textrm{Lambek}^D$}
\newcommand{\lnld}{$\textrm{LNL}_D$}

@@

\section{Elided Syntax}

\max{TODO: add prose}

@@

>>>>

\newcommand{\Subst}{\textrm{Subst}}
\newcommand{\SPF}{\mathsf{SPF}}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\K}{\mathsf{K}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\roll}{\mathsf{roll}}
\newcommand{\fold}{\mathsf{fold}}
\newcommand{\inl}{\mathsf{inl}}
\newcommand{\inr}{\mathsf{inr}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\String}{\textbf{String}}
\newcommand{\stringg}{\mathtt{String}}
\newcommand{\charg}{\mathtt{Char}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Syn}{\mathbf{Synx}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathtt{Bool}}
\newcommand{\true}{\mathtt{true}}
\newcommand{\false}{\mathtt{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
\newcommand{\theoryabbv}{$\textrm{Lambek}^D$}
\newcommand{\lnld}{$\textrm{LNL}_D$}

@@

\section{Syntax}

In this section we include the elided syntactic forms, as well as
definitions and basic properties of linear and non-linear
substitution.

@@

\begin{definition}
  The set of (non-linear) substitutions $\gamma \in
  \textrm{Subst}(\Gamma,\Gamma')$ where $\Gamma \isCtx$ and $\Gamma'
  \isCtx$ is defined by recursion on $\Gamma$:
  \begin{align*}
    \textrm{Subst}(\Gamma,\cdot) &= \{\cdot \}\\
    \textrm{Subst}(\Gamma,\Gamma',x:A) &= \{ (\gamma,M/x) \pipe \gamma \in \textrm{Subst}(\Gamma,\Gamma') \wedge \Gamma \vdash M : A[\gamma] \}
  \end{align*}
  simultaneously with an action of substitution on types, terms,
  etc. in the standard way.

  By induction we can define an identity substitution $\textrm{id} :
  \textrm{Subst}(\Gamma,\Gamma)$ which maps each variable to itself,
  and composition of substitutions
  \begin{align*}
    \cdot \circ \gamma' &= \cdot\\
    (\gamma,M/x) \circ \gamma' &= (\gamma\circ \gamma') , M[\gamma']/x
  \end{align*}

  It is straightforward, but laborious to establish that all forms in
  the type theory that are parameterized by a $\Gamma$ support an
  admissible action of substitution:
  \begin{mathpar}
    \inferrule
    {\Gamma \vdash M : X \and \gamma \in \Subst(\Gamma',\Gamma)}
    {\Gamma' \vdash M[\gamma] : X[\gamma]}

    \inferrule
    {\Gamma;\Delta \vdash e : A \and \gamma \in \Subst(\Gamma',\Gamma)}
    {\Gamma'; \Delta[\gamma] \vdash e[\gamma] : A[\gamma]}
  \end{mathpar}
  et cetera.
\end{definition}

\begin{definition}
  Let $\Gamma \vdash \Delta \isLinCtx$ and $\Gamma \vdash \Delta'
  \isLinCtx$. The set of linear substitutions $\Subst(\Delta',\Delta)$
  is defined by recursion on $\Delta$:
  \begin{align*}
    \Subst(\Delta',\cdot) &= \{ \cdot \pipe \Delta' = \cdot \}\\
    \Subst(\Delta',(\Delta,a:A)) &= \{ (\delta, e/a) \pipe \delta \in \Subst(\Delta_1, \Delta) \and \Delta_2 \vdash e : A \and \Delta' = \Delta_1,\Delta_2\}
  \end{align*}

  Given substitutions $\delta_1 \in \Subst(\Delta_1', \Delta_1)$ and
  $\delta_2 \in \Subst(\Delta_2', \Delta_2)$, we can define a
  substitution $\delta_1,\delta_2 \in
  \Subst((\Delta_1',\Delta_2'),(\Delta_1,\Delta_2))$. Furthermore, for
  any substitution $\delta \in \Subst(\Delta,(\Delta_1,\Delta_2))$, we
  can deconstruct $\delta = \delta_1,\delta_2$ with $\delta_1 \in
  \Subst(\Delta_1', \Delta_1)$ and $\delta_2 \in \Subst(\Delta_2',
  \Delta_2)$.
  %% Again for any $\Gamma \vdash \Delta$, we can define an identity
  %% substitution mapping each variable to itself. Additionally we can define composition of 
\end{definition}

\begin{definition}
  Given any $\Gamma ; \Delta \vdash e : A$ and $\delta \in
  \Subst(\Delta',\Delta)$, we define the substitution $\Gamma; \Delta
  \vdash e[\delta] : A$ as follows, frequently using the inversion
  principle to split the substitution into constituent components.
  \begin{align*}
    a[e/a] &= e\\
    (e_1,e_2)[\delta_1,\delta_2] &= (e_1[\delta_1], e_2[\delta_2])\\
    (\letin {(a , b)} e {e'})[\delta_1,\delta_2,\delta_3] &= \letin{(a,b)} {e[\delta_2]}{e'[\delta_1,a/a,b/b,\delta_2]}\\
    ()[\cdot] &= ()\\
    \letin{()} e {e'}[\delta_1,\delta_2,\delta_3] &= \letin{()} {e[\delta_2]}{e'[\delta_1,a/a,b/b,\delta_2]}\\
    (\lamblto {a} e)[\delta] &= \lamblto a {e[\delta,a/a]}\\
    (\applto {e'} {e})[\delta_1,\delta_2] &= \applto {e'[\delta_1]} {e[\delta_2]}\\
    (\lamblto {a} e)[\delta] &= \lamblto a {e[\delta,a/a]}\\
    (\applto {e'} {e})[\delta_1,\delta_2] &= \applto {e'[\delta_1]} {e[\delta_2]}\\
    (\lambtol {a} e)[\delta] &= \lamblto a {e[a/a,\delta]}\\
    (\apptol {e'} {e})[\delta_1,\delta_2] &= \apptol {e'[\delta_1]} {e[\delta_2]}\\
    (\dlamb x e)[\delta] &= \dlamb x {e[\delta]}\\
    (e\,.\pi\,M)[\delta] &= (e[\delta]\,.\pi\,M)\\
    (\sigma\,M\,e)[\delta] &= \sigma\,M\,e[\delta]\\
    (\letin{\sigma\,x\,a}{e}{e'})[\delta_1,\delta_2,\delta_3] &= \letin{\sigma\,x\,a} {e[\delta_2]}{e'[\delta_1,\delta_3]}\\
    (\equalizerin{e})[\delta] &= \equalizerin{e[\delta]}\\
    (\equalizerpi{e})[\delta] &= \equalizerpi{e[\delta]}
  \end{align*}
\end{definition}

\section{Denotational Semantics}

Next we define the term semantics for linear terms.

\begin{definition}
  For any $\Gamma \vdash \Delta_1,\Delta_2 \isLinCtx$ and $\gamma \in
  \sem{\Gamma}$ there is an isomorphism $m_{\Delta_1,\Delta_2} :
  \sem{\Delta_1, \Delta_2}\gamma \cong \sem{\Delta_1}\gamma \otimes
  \sem{\Delta_2}\gamma$.

  This can be extended to a sequence of contexts of any length.
\end{definition}

\begin{mathpar}
  \inferrule
  {\Delta_1 \vdash e_1 : A_1 \and \Delta_2 \vdash e_2 : A_2}
  {\sem{(e_1,e_2)}\gamma\,w\,\delta
    = \letin{(w_1,w_2,\delta_1,\delta_2)} {m(\Delta_1,\Delta_2)(\delta)} {(w_1,w_2,\sem{e_1}\gamma\,w_1\,\delta_1, \sem{e_2}\gamma\,w_2\,\delta_2)}
  }

  \inferrule
  {\Delta_2 \vdash e : A \otimes B \and \Delta_1,a:A,b:B,\Delta_3 \vdash e' : C}
  {\sem{\letin{(a,b)}{e}{e'}}\gamma\,w\,\delta =
    \letin{(w_1,w_2w_3,\delta_1,(w_2,w_3,\delta_2,\delta_3))}{m_{\Delta_1,\Delta_2,\Delta_3}(\delta)}
    \letin{(w_{2,1},w_{2,2},a,b)}{\sem{e}\gamma\,w_2\,\delta_2}
    {\sem{e'}\gamma\,w\,(w_1,w_{2,1}w_{2,2}w_3, \delta_1, (w_{2,1},w_{2,2}w_3, a, (w_{2,2},w_3, b, \delta_3)))}
    }
\end{mathpar}
\begin{align*}
  \sem{(e_1,e_2)}\gamma\,w\,\delta &=
  \letin{(w_1,w_2,\delta_1,\delta_2)} {m(\Delta_1,\Delta_2)(\delta)} {(w_1,w_2,\sem{e_1}\gamma\,w_1\,\delta_1, \sem{e_2}\gamma\,w_2\,\delta_2)}\\
\end{align*}
